\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-02-21 analysis/paper/paper.tex

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with a description
of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression.
\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

@@

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

@@

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}.
\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}

@@

\begin{CodeInput}
R> data(support)
R> support_popTime <- popTime(support,
R+                            time = "d.time", event = "death")
R> plot(support_popTime)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{../figures/support-poptime-1}

\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

Before we proceed with the analysis, we first remove hospital death as a
covariate, as this is directly informative of death and cannot be used
as a predictor. Then we select one observation at random. We will
estimate the cumulative incidence curve for this individual. The other
observations will be used to fit the hazard function.

@@

\begin{CodeInput}
R> support <- dplyr::select(support, -hospdead)
R> 
R> # Randomly select one observation
R> indiv <- sample(1:nrow(support), size = 1)
\end{CodeInput}
\end{CodeChunk}

Our main objective is to estimate the cumulative incidence of death for
a given covariate profile. Given the number of covariates available, we
will do this using regularized estimation. As \pkg{casebase} makes use
of the \pkg{glmnet} package, we will use the function
\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
Specifically, we input a matrix \code{y}, which contains the time and
event variables, and a matrix \code{x}, which contains all other
variables we would like to include in the model. For the SUPPORT
dataset, all factor variables were converted into dummy variables.

@@

\begin{CodeInput}
R> # Create design matrix without the time variable
R> support_matrix <- model.matrix(death ~ . - d.time, 
R+                                data = support)
R> # Remove intercept term
R> support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
R> 
R> # Create test and train dataset
R> support_test <- support_matrix[indiv,,drop = FALSE]
R> 
R> support_train <- support_matrix[-indiv,]
R> y <- as.matrix(support[-indiv, c("death", "d.time")])
\end{CodeInput}
\end{CodeChunk}

We are now ready to estimate the hazard function. We use lasso
penalization; this is achieved by setting \code{alpha} to 1. Also, note
that the argument \code{formula_time} can be used to specify the
function form for time; in our example, we model time linearly.

@@

\begin{CodeInput}
R> support_cb <- fitSmoothHazard.fit(support_train, y, 
R+                                   formula_time = ~ d.time, 
R+                                   family = "glmnet", 
R+                                   event = "death", time = "d.time", 
R+                                   alpha = 1, ratio = 100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model object and use the \code{absoluteRisk} function
to compute our estimate of the cumulative incidence function. Since
\code{glmnet} fitted the penalized logistic regression for a range of
penalty terms, we need to specify which one we want to use for our
prediction. We select \code{s = "lambda.1se"}, as recommended.

@@

\begin{CodeInput}
R> time_points <- seq(1, max(support$d.time), by = 0.5)
R> support_risk <- absoluteRisk(support_cb,
R+                              time = time_points,
R+                              newdata = support_test,
R+                              s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of Cox regression,
and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
can fit the Cox regression model using regularized estimation.

@@

\begin{CodeInput}
R> # Create survival object for glmnet
R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
R+                               event = y[, "death"])
R> 
R> support_cox <- glmnet::cv.glmnet(x = support_train, 
R+                                  y = surv_glmnet,
R+                                  family = "cox", alpha = 1)
\end{CodeInput}
\end{CodeChunk}

However, \pkg{glmnet} does not provide any functionality for estimating
the baseline hazard, and the functionalities in the \pkg{survival}
package do not work on the output from \code{cv.glmnet}. To circumvent
this issue and get an estimate of the baseline hazard, we followed the
following steps. First, we fitted a non-regularized Cox model using only
the variables selected by \pkg{glmnet}; the baseline hazard will be
estimated using this object. Then, we replaced the estimates of the
regression coefficients by the estimates from \pkg{glmnet}. Finally,
this modified \code{coxph} object is fed to the \code{survival::survfit}
function to obtain the estimated survival function.

@@

\begin{CodeInput}
R> # Identifying selected covariates
R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
R+                                   s = "lambda.1se")
R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
R> # Creating a new dataset that only contains the covariates selected by glmnet
R> support_cox_data <- as.data.frame(cbind(y,
R+                                         support_train[, nonzero_covariates_cox$X1]))
R> # Fitting a cox model
R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
R+                           data = support_cox_data)
R> # Replace the estimate with those obtained by glmnet
R> support_cox$coefficients <- nonzero_coef_cox@x
R> 
R> # Fitting CI curve for cox+glmnet
R> support_cox_risk <- survival::survfit(support_cox,
R+                                       newdata = as.data.frame(support_test),
R+                                       time = time_points,
R+                                       type = "breslow")
\end{CodeInput}
\end{CodeChunk}

We used the \pkg{survival} package to calculate the Kaplan-Meier
estimate of the survival function.

@@

\begin{CodeInput}
R> km <- survival::Surv(time = support$d.time, 
R+                      event = support$death)
R> abKm <- survival::survfit(km ~ 1,
R+                           type = 'kaplan-meier')
\end{CodeInput}
\end{CodeChunk}

Now that our three cumulative incidence functions have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
\pkg{casebase} have a lower estimated cumulative incidence by the end of
the study, when compared to Kaplan-Meier.
\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}

{\centering \includegraphics{../figures/abSupportComparison-1}

\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
\end{figure}
\end{CodeChunk}

@@

In the previous case studies, we only considered covariates that were
fixed at baseline. In this next case study, we will use the Stanford
Heart Transplant data
\citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
case-base sampling can also be used in the context of time-varying
covariates. As an example that already appeared in the literature,
case-base sampling was to study vaccination safety, where the exposure
period was defined as the week following vaccination
\citep{saarela2015case}. Hence, the main covariate of interest,
i.e.~exposure to the vaccine, was changing over time. In this context,
case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    795    760    729    724
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.27
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.19
\end{CodeOutput}
\end{CodeChunk}

@@

This report was generated on 2020-02-07 11:07:26 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 3.6.2 (2019-12-12)
#> Platform: x86_64-pc-linux-gnu (64-bit)
#> Running under: Ubuntu 18.04.4 LTS
#> 
#> Matrix products: default
#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so
#> 
#> locale:
#>  [1] LC_CTYPE=en_CA.UTF-8       LC_NUMERIC=C              
#>  [3] LC_TIME=en_CA.UTF-8        LC_COLLATE=en_CA.UTF-8    
#>  [5] LC_MONETARY=en_CA.UTF-8    LC_MESSAGES=en_CA.UTF-8   
#>  [7] LC_PAPER=en_CA.UTF-8       LC_NAME=C                 
#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
#> [11] LC_MEASUREMENT=en_CA.UTF-8 LC_IDENTIFICATION=C       
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] lubridate_1.7.4     timereg_1.9.4       glue_1.3.1         
#>  [4] cowplot_1.0.0       pracma_2.2.9        kableExtra_1.1.0   
#>  [7] glmnet_3.0-1        Matrix_1.2-18       casebase_0.2.1.9001
#> [10] survival_3.1-8      magrittr_1.5        forcats_0.4.0      
#> [13] stringr_1.4.0       dplyr_0.8.3         purrr_0.3.3        
#> [16] readr_1.3.1         tidyr_1.0.0         tibble_2.1.3       
#> [19] ggplot2_3.2.1       tidyverse_1.3.0    
#> 
#> loaded via a namespace (and not attached):
#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6       
#>  [4] viridisLite_0.3.0   foreach_1.4.7       modelr_0.1.5       
#>  [7] assertthat_0.2.1    stats4_3.6.2        cellranger_1.1.0   
#> [10] yaml_2.2.0          numDeriv_2016.8-1.1 pillar_1.4.3       
#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.23      
#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.5         
#> [19] htmltools_0.4.0     pkgconfig_2.0.3     broom_0.5.2        
#> [22] haven_2.2.0         scales_1.1.0        webshot_0.5.2      
#> [25] lava_1.6.6          mgcv_1.8-31         rticles_0.12       
#> [28] generics_0.0.2      farver_2.0.1        ellipsis_0.3.0     
#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.1          
#> [34] crayon_1.3.4        readxl_1.3.1        evaluate_0.14      
#> [37] fs_1.3.1            fansi_0.4.1         MASS_7.3-51.5      
#> [40] nlme_3.1-143        xml2_1.2.2          tools_3.6.2        
#> [43] data.table_1.12.8   hms_0.5.2           lifecycle_0.1.0    
#> [46] munsell_0.5.0       reprex_0.3.0        compiler_3.6.2     
#> [49] rlang_0.4.2         grid_3.6.2          iterators_1.0.12   
#> [52] rstudioapi_0.10     labeling_0.3        rmarkdown_1.18     
#> [55] gtable_0.3.0        codetools_0.2-16    DBI_1.0.0          
#> [58] reshape2_1.4.3      R6_2.4.1            knitr_1.26         
#> [61] zeallot_0.1.0       shape_1.4.4         stringi_1.4.4      
#> [64] Rcpp_1.0.3          vctrs_0.2.1         dbplyr_1.4.2       
#> [67] tidyselect_0.2.5    xfun_0.11
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
#> Head:     [f1e32e8] 2020-02-07: add discussion of flexible sampling
\end{CodeOutput}
\end{CodeChunk}

>>>>

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with some of the
functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation for ridge.
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
are the only packages to offer this functionality.
\textcolor{red}{I can dive deeper to determine if this is a fair statement}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Several \proglang{R} packages implement methodologies for competing risk
analysis; for a different perspective, see Mahani \& Sharabiani
\citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk} provides
methods for cause-specific subdistributions, such as in the Fine-Gray
model \citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific cumulative incidence functions from
unadjusted, non-parametric survival functions. Our package
\pkg{casebase} also provides functionalities for competing risk analysis
by estimating the cause-specific hazards, and from these quantities, we
can derive the cause-specific cumulative incidence functions.

@@

\caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

@@

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}.

@@

\begin{CodeInput}
R> data(support)
R> support_popTime <- popTime(support,
R>                            time = "d.time", event = "death",percentile_number = 0.31)
R> plot(support_popTime)
\end{CodeInput}
\end{CodeChunk}

Before we proceed with the analysis, we first remove hospital death as a
covariate, as this is directly informative of death. All covariates will
be used to estimate the hazard function. The hazard function will be
used to estimate the cumulative incidence of each individual in the
study. Then, the average cumulative incidience curve for all
observations will be used for our comparison.

Our main objective is to compute the absolute risk of death for a given
set of covariates, using regularization when fitting our model. First,
we fit a smooth hazard to the data; for the sake of this example, we
opted for a linear term for time. As casebase makes use of the glmnet
package, we interact with the fitsmoothhazard.fit function using a
matrix interface, where y contains the time and event variables and x
contains all other variables we would like to include in the model. For
the SUPPORT dataset, all factor variables were converted into dummy
variables. Here, we used lasso penalization by setting alpha to 1.

@@

\begin{CodeInput}
R> xInteractions=x*y[,2]
R> colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
R> xFull=cbind(x,xInteractions)
R> cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
R>                                        formula_time = ~d.time,alpha=1,ratio=rat)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model and use the absoluteRisk function to integrate
and retrieve our absolute risk function.

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
R> 
R> cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The cox regression absolute risk
curve with lasso penalization required two extra steps to retrieve.
Coxnet from the glmnet package has tools to fit a regularized cox model,
but requires a survival object skeleton of the selected variables, so
that the survival package tools can be used to retrive the absolute
risk. We handle this by fitting a regularized cox model with the glmnet
package, and a cox model from the survival package with the selected
variables from the regularized fit. This second model serves as a
skeleton that permits the use of survival package functions. The
coefficients from the regularized model will replace the coefficients in
our skeleton. the resulting model will have the correct coefficients
when integrating, but will have an incorrect standard error. For the
purposes of this case study, we are only interested in the absolute risk
curve itself and not the standard error.

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> 
R> xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
R> colnames(xCoxI)[1]="(Intercept)"
R> 
R> #hazard for cox using glmnet
R> coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> #plot(coxNet)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
R> #The coefficients of this object will be replaced with the estimates from coxNet.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxNetFitI$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
R> #posCox=nonzero_covariate_cox$X1
R> #newDataCoxI=cleanCoxDataI[,posCox]
R> 
R> abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
R> 
R> abCNI<-abcoxNetI
R> abCNI$surv=rowMeans(abCNI$surv)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
R> coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
R> abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
R> abCFI<-abcoxFitI
R> abCFI$surv<-rowMeans(abCFI$surv)
R> #cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
R>        #                                ratio=rat,data=fullDataCox)
R> #summary(cb.ModelNorm)#need to figure out how to deal with NAs
R> #casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
R> #cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
R> 
\end{CodeInput}
\end{CodeChunk}

We used the survival package to calculate the kaplan-meier absolute risk
function.

@@

\begin{CodeInput}
R> 
R> coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
R> abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
R> 
R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
R> # (Kaplan-Meier risk curve)
R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
\end{CodeInput}
\end{CodeChunk}

Now that our four cumulative incidence estimations have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
decrease the absolute risk by the end of the study, in comparison to
kaplan-meier.

@@

\begin{CodeInput}
R> 
R> #get covariates excluding d.time
R> estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
R> estimatescb$Model="casebase"
R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
R> estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
R> estimatescox$Model="Cox"
R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
R> estimates=rbind(estimatescb, estimatescox)
R> estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
R> library(ggstance)# for vertical dodge
R>   ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
R>         geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
R>         geom_point(position=ggstance::position_dodgev(height=0.2))
R> 
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #get covariates excluding d.time
R> library(data.table)
R> estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
R> estimatescb$Model="casebase"
R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
R> estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
R> estimatescox$Model="Cox"
R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
R> estimatesI=rbind(estimatescb, estimatescox)
R> estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
R> library(ggstance)# for vertical dodge
R>   ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
R>         geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
R>         geom_point(position = position_dodge(width = 0.5))
R>   
R>     
\end{CodeInput}
\end{CodeChunk}

@@

In the previous case studies, we only considered covariates that were
fixed at baseline. In this next case study, we will use the Stanford
Heart Transplant data
\citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
case-base sampling can also be used in the context of time-varying
covariates. As an example that already appeared in the literature,
case-base sampling was used to study vaccination safety, where the
exposure period was defined as the week following vaccination
\citep{saarela2015case}. Hence, the main covariate of interest,
i.e.~exposure to the vaccine, was changing over time. In this context,
case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    791    760    727    724
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.28
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.18
\end{CodeOutput}
\end{CodeChunk}

@@

This report was generated on 2020-02-21 13:47:19 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 3.6.2 (2019-12-12)
#> Platform: x86_64-pc-linux-gnu (64-bit)
#> Running under: Ubuntu 18.04.4 LTS
#> 
#> Matrix products: default
#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so
#> 
#> locale:
#>  [1] LC_CTYPE=en_CA.UTF-8       LC_NUMERIC=C              
#>  [3] LC_TIME=en_CA.UTF-8        LC_COLLATE=en_CA.UTF-8    
#>  [5] LC_MONETARY=en_CA.UTF-8    LC_MESSAGES=en_CA.UTF-8   
#>  [7] LC_PAPER=en_CA.UTF-8       LC_NAME=C                 
#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
#> [11] LC_MEASUREMENT=en_CA.UTF-8 LC_IDENTIFICATION=C       
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] lubridate_1.7.4     timereg_1.9.4       glue_1.3.1         
#>  [4] cowplot_1.0.0       pracma_2.2.9        kableExtra_1.1.0   
#>  [7] data.table_1.12.8   glmnet_3.0-1        Matrix_1.2-18      
#> [10] casebase_0.2.1.9001 survival_3.1-8      magrittr_1.5       
#> [13] forcats_0.4.0       stringr_1.4.0       dplyr_0.8.3        
#> [16] purrr_0.3.3         readr_1.3.1         tidyr_1.0.0        
#> [19] tibble_2.1.3        ggplot2_3.2.1       tidyverse_1.3.0    
#> 
#> loaded via a namespace (and not attached):
#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6       
#>  [4] viridisLite_0.3.0   foreach_1.4.7       modelr_0.1.5       
#>  [7] assertthat_0.2.1    stats4_3.6.2        cellranger_1.1.0   
#> [10] yaml_2.2.0          numDeriv_2016.8-1.1 pillar_1.4.3       
#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.24      
#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.5         
#> [19] htmltools_0.4.0     pkgconfig_2.0.3     broom_0.5.2        
#> [22] haven_2.2.0         scales_1.1.0        webshot_0.5.2      
#> [25] lava_1.6.6          mgcv_1.8-31         rticles_0.12       
#> [28] generics_0.0.2      farver_2.0.3        ellipsis_0.3.0     
#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.1          
#> [34] crayon_1.3.4        readxl_1.3.1        evaluate_0.14      
#> [37] fs_1.3.1            fansi_0.4.1         MASS_7.3-51.5      
#> [40] nlme_3.1-144        xml2_1.2.2          tools_3.6.2        
#> [43] hms_0.5.2           lifecycle_0.1.0     munsell_0.5.0      
#> [46] reprex_0.3.0        compiler_3.6.2      rlang_0.4.4        
#> [49] grid_3.6.2          iterators_1.0.12    rstudioapi_0.10    
#> [52] labeling_0.3        rmarkdown_1.18      gtable_0.3.0       
#> [55] codetools_0.2-16    DBI_1.0.0           reshape2_1.4.3     
#> [58] R6_2.4.1            knitr_1.26          shape_1.4.4        
#> [61] stringi_1.4.6       Rcpp_1.0.3          vctrs_0.2.3        
#> [64] dbplyr_1.4.2        tidyselect_0.2.5    xfun_0.11
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    Jesse-Islam-master /home/turgeonm/Documents/git_repos/cbpaper
#> Head:     [3ec29a1] 2020-02-07: Merge pull request #18 from sahirbhatnagar/intro
\end{CodeOutput}
\end{CodeChunk}
